### 2016-03-16 hans/bayes-tardis-luminosity-prediction.tex
[draft] Polish description of Gamma model

\DeclareMathOperator{\GammaDist}{Gamma}
\newcommand{\npack}{n_p}
\newcommand{\lmax}{\ell_{\rm max}}
\newcommand{\Lmax}{L_{\rm max}}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential

@@

The Gamma distribution  with shape parameter $\alpha$ and rate
parameter $\beta$ is defined as
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
\end{align}
The support of $\GammaDist(x \cond \alpha, \beta)$ is $[0, \infty]$
whereas the luminosity of real packets is bounded from above and has a
peak close to but not at the boundary. \fred{refer to plot of real
  luminosity packets in some frequency bin} In order to describe the
luminosity distribution of real packets through $\GammaDist(\alpha,
\beta)$, we have to transform the luminosities. To this end, we first
take the largest luminosity of all TARDIS packets across the entire
frequency range and increase it slightly to avoid numerical
difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
\end{align}
where $ \epsilon=10^{-6}$.  We assume the total number of samples
across all bins is large, hence the effect of considering $\lmax$ a
constant independent of the data is negligible. We then transform the
samples $\bml \to \bmx$ as
\begin{align}
  \lleq{gmf}
  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
\end{align}
and similarly for $L \to X$, then we infer $P(X \cond \nu,  \npack, \bmx,
\bmnu)$. Only in the end, we transform back to a probability density
in $L$ via
\begin{align}
  \lleq{gmg}
  p(L  \cond \nu,  \npack, \bml, \bmnu) = p(X  \cond \nu,  \npack, \bmx, \bmnu) \frac{1}{\lmax}
\end{align}
\hans{So far, $m$ appears only in your todo comment above. I would
  suggest a short discussion of this already in Section
  \ref{sec:dfnr}. We could then simply rewrite all equations below in
  terms of $m$ rather than $b$.} %
\fred{I removed $m$ whereever I saw it} We observe good agreement
between the best fit of a Gamma distribution to TARDIS packet
luminosities; an example is shown shown in \reffig{gamma}.  While not
obvious at first, a major numerical simplification arises because of
the $L \to X$ transformation. Rewriting the master
equation~(\ref{pmp}) formally as
\begin{align}
  \lleq{gmh}
  p(L  \cond \nu,  \npack, \bml, \bmnu) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
\nu, \npack, \bml, \bmnu)
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single luminosity, $p(L \cond N=1,
\nu, \bml, \bmnu)$ is narrow in the sense that the standard deviation
relative to the mean is at the percent level. This implies that $p(L
\cond \nu, \npack, \bml, \bmnu)$ has one mode from each $N$ and all
modes are clearly separated. The transformation $L \to x$ effects a
strong overlap between the distributions $p(X \cond N, \nu, \npack, \bmx, \bmnu)$
with similar $N$, so $p(X \cond \nu, \npack, \bmx, \bmnu)$ can be well
approximated by a unimodal distribution; see \reffig{hedgehog}.
\fred{I'm not 100 \% sure but $p(X \cond \nu, \npack, \bmx, \bmnu)$ should still
  have $N \to \infty$ modes but they aren't visible}
\missingfigure{from hedgehog to single peak \label{fig:hedgehog}}

@@

For our purposes, the most useful property of the Gamma
distribution is infinite divisibility; i.e., the sum of $N$ Gamma
variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
  \boxed{
    X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
    }
\end{align}
With
\begin{align}
  \lleq{gmi}
  \alpha, \beta &= \alpha(\nu \cond \bmphi), \beta(\nu \cond \bmphi),\\
  p(X_j \cond \bmphi, \nu) &=
  \GammaDist(X_j \cond \alpha, \beta),\\
  p(\bmX \cond N, \bmphi, \nu) &= \prod_{j=1}^N p(X_j \cond \bmphi, \nu)
\end{align}
 \refeq{gmb} is equivalent to
\begin{align}
  \lleq{gmc}
 p(X \cond N, \bmphi, \nu) &= \int \rmdx{\bmX} \delta(X - \textstyle\sum_j X_j)
  \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)
\end{align}
and this implies that the master equation~(\ref{pmp}) considerably simplifies as
\begin{align}
  \lleq{gmd}
    p(X\cond \nu, \npack,\bmx,\bmnu)
  = \frac{1}{p(\bmx\cond \bmnu)}
  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi, \nu)\, p(\bmx\cond \bmphi, \npack, \bmnu)
  \, p(\bmphi).
\end{align}

\subsection*{Approximations}

To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
two difficulties: the sum over $N$ and the integral over
$\bmphi$. Regarding the latter, there is a conjugate prior but
unfortunately the normalization constant cannot be evaluated in closed
form. This conjugate prior is thus of no use for us, and we turn to a
noninformative prior. \fred{Be explicit about prior choice} The
authors of \cite{moala2013bayesian} present an overview of possible
choices and conclude that the point estimate given by the mode of the
posterior is nearly insensitive to the prior choice for 30 samples in
the iid case. Translated to our problem, this would be 30 packets at
the same frequency. From TARDIS, we only get packets at different
frequencies. In the following, let $m$ denote the effective number of
packets close to $\nu$ where ``close'' is deliberately left
unspecified.  If $m$ is large, we can well determine $\phi$
and thus $p(X_j \cond N, \bmphi, \nu)$.

@@

The following approximations can be used in different regions of the
parameter space spanned by $n$ and $m$. %
\hans{$n$ is TARDIS output ``data'', so not a parameter. From the
  above paragraph, $m$ seems to be a parameter governing rebinning to
  nonuniform bins. I would have no idea how to handle $m$ formally in
  a Bayesian way. If $m{=}150$ is the result of optimising the fitting
  stability, then presumably it will be kept constant. In that case,
  it would be treated like a constant rather than a parameter in the
  usual sense.}
\fred{Yes, $n$ is fixed from TARDIS output. $m$ is defined only for this paragraph}
\fred{Update approximations: don't need moments any longer}
\begin{mnumerate}
  \item \label{large-m} For large $m$, the posterior $p(\bmphi \cond
  \npack, \bmx, \bmnu) = p(\bmx\cond \bmphi, \npack, \bmnu) \, p(\bmphi) / p(\bmx \cond \npack, \bmnu)$
  approaches a multivariate Gaussian. We can then compute the various
  moments $E_\bmphi[f(x)]$ with $f(x)=1,x,x^2$ by Laplace's method to
  good accuracy.
  \item \label{very-large-m} In the limit $m \to \infty$, the
  covariance of the posterior $p(\bmphi \cond \npack, \bmx, \bmnu)$
  tends to zero as $1/m$ and we can then fix the parameters $\bmphi$
  at the posterior mode $\bmphi^*$; i.e., $p(\bmphi \cond \npack,
  \bmx, \bmnu) \to \delta(\bmphi - \bmphi^*)$.
  \item \label{large-n} For $n$ large, the cumulative of the Negative
  Binomial $p(N \cond n)$ can be approximated by a Gaussian, defining
  a region around $N=n$ with the dominant contribution to the sum over
  $N$ in \refeq{gmd}. The contribution around $N=0$ is negligible, and
  we can replace $\sum_n \to \int \rmdx{N}$. The density $p(N \cond n) P(X
  \cond N,m, \bmx)$ is close to a Gaussian and the integral can be
  accurately approximated by Laplace's method which requires the
  Hessian and the mode. We benefit from partial error cancelation
  because we apply Laplace's method both in the numerator and
  denominator~\cite{tierney_accurate_1986} as each moment has to be
  normalized by the evidence $p(\bmx \cond m)$. The former is
  available in closed form and the latter has to be determined
  numerically. Fortunately, Newton's method leads to a solution in
  only a few steps~\cite{minka2002estimating}. \fred{The reference is
    for max. likelihood of a $\GammaDist(\cdot)$ but I'm sure the
    performance is similar for max. posterior and the first moments}
  \item \label{central-limit} The central limit theorem states that for large $N$,
  $P(X \cond N, \bmphi) \approx \mathcal{N}(X \cond N \,
  E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For $m$ large enough, we could
  then plug in the sample-mean and sample-variance estimates and
  bypass the assumption of a Gamma distribution altogether.
  \item \label{large-n-large-m} For large $n$ and $m$, $ p(X\cond
  n,\bmx,m)$ can be approximated by a Gaussian whose mean and variance
  are given by \refeq{mgo} and \refeq{mgr} in terms of
  $\bmphi$-expectation values. The latter are integrals with no known
  closed-form solutions but we can easily and accurately approximate
  them with Laplace's method.
\end{mnumerate}
To simplify, we assume that $m$ is large. In our application, we may
have to combine several frequency bins to achieve $m \gtrsim 150$ for
the purpose of infering $\bmphi$. This is equivalent to assuming that
the distribution changes only slowly with the frequency, a condition
that we verified for TARDIS. It is thus justified to use
approximations \ref{large-m} or even \ref{very-large-m}.

@@

\subsection*{Comparison of approximations}

>>>>

\DeclareMathOperator{\GammaDist}{Gamma}
\newcommand{\Kalpha}{{K_\alpha}}
\newcommand{\Kbeta}{{K_\beta}}
\newcommand{\npack}{n_p}
\newcommand{\lmax}{\ell_{\rm max}}
\newcommand{\Lmax}{L_{\rm max}}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential

@@

The Gamma distribution  with shape parameter $\alpha$ and rate
parameter $\beta$ is defined as
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
\end{align}
The support of $\GammaDist(x \cond \alpha, \beta)$ is $[0, \infty]$
whereas the luminosity of real packets is bounded from above and has a
peak close to but not at the boundary. \fred{refer to plot of real
  luminosity packets in some frequency bin} In order to describe the
luminosity distribution of real packets through $\GammaDist(\alpha,
\beta)$, we have to transform the luminosities. To this end, we first
take the largest luminosity of all TARDIS packets across the entire
frequency range and increase it slightly to avoid numerical
difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
\end{align}
where $ \epsilon=10^{-6}$.  We assume the total number of samples
across all bins is large, hence the effect of considering $\lmax$ a
constant independent of the data is negligible. We then transform the
samples $\bml \to \bmx$ as
\begin{align}
  \lleq{gmf}
  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
\end{align}
and similarly for $L \to X$, then we infer $P(X \cond \nu,  \npack, \bmx,
\bmnu)$. Only in the end, we transform back to a probability density
in $L$ via
\begin{align}
  \lleq{gmg}
  p(L  \cond \nu,  \npack, \bml, \bmnu) = p(X  \cond \nu,  \npack, \bmx, \bmnu) \frac{1}{\lmax}
\end{align}
\hans{So far, $m$ appears only in your todo comment above. I would
  suggest a short discussion of this already in Section
  \ref{sec:dfnr}. We could then simply rewrite all equations below in
  terms of $m$ rather than $b$.} %
\fred{I removed $m$ whereever I saw it except below where I define it
  explicitly}
%
We observe good agreement between the best fit of a Gamma distribution
to TARDIS packet luminosities; an example is shown shown in
\reffig{gamma}.  While not obvious at first, a major numerical
simplification arises because of the $L \to X$
transformation. Rewriting the master equation~(\ref{pmp}) formally as
\begin{align}
  \lleq{gmh}
  p(L  \cond \nu,  \npack, \bml, \bmnu) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
\nu, \npack, \bml, \bmnu)
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single luminosity, $p(L \cond N=1,
\nu, \bml, \bmnu)$ is narrow in the sense that the standard deviation
relative to the mean is at the percent level. This implies that $p(L
\cond \nu, \npack, \bml, \bmnu)$ has one mode from each $N$ and all
modes are clearly separated. The transformation $L \to x$ effects a
strong overlap between the distributions $p(X \cond N, \nu, \npack, \bmx, \bmnu)$
with similar $N$, so $p(X \cond \nu, \npack, \bmx, \bmnu)$ can be well
approximated by a unimodal distribution; see \reffig{hedgehog}.
\fred{I'm not 100 \% sure but $p(X \cond \nu, \npack, \bmx, \bmnu)$ should still
  have $N \to \infty$ modes but they aren't visible}
\missingfigure{from hedgehog to single peak \label{fig:hedgehog}}

@@

For our purposes, the most useful property of the Gamma
distribution is infinite divisibility; i.e., the sum of $N$ Gamma
variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
    X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
\end{align}
Let us assume that $\alpha$ and $\beta$ are functions of $\nu$ with unknown parameters $\bmphi$
\begin{align}
  \lleq{gmi}
  \alpha &= \alpha(\nu \cond \bmphi),\\
  \beta &= \beta(\nu \cond \bmphi),\\
  p(X_j \cond \bmphi, \nu) &=
  \GammaDist(X_j \cond \alpha, \beta),\\
  p(\bmX \cond N, \bmphi, \nu) &= \prod_{j=1}^N p(X_j \cond \bmphi, \nu).
\end{align}
With these definitions, \refeq{gmb} is equivalent to
\begin{align}
  \lleq{gmc}
 p(X \cond N, \bmphi, \nu) &= \int \rmdx{\bmX} \delta(X - \textstyle\sum_j X_j)
  \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)
\end{align}
and this implies that the master equation~(\ref{pmp}) considerably simplifies as
\begin{align}
  \lleq{gmd}
    \boxed{
    p(X\cond \nu, \npack,\bmx,\bmnu)
  = \frac{1}{p(\bmx\cond \bmnu)}
  \sum_N p(N\cond n)\, \int \rmdx{\bmphi} \GammaDist(X_j \cond \alpha, \beta) % p(X \cond N, \bmphi, \nu)
  \, p(\bmx\cond \bmphi, \npack, \bmnu)
  \, p(\bmphi).
  }
\end{align}

\subsubsection*{Linear model}

In the absence of a physical model to determine $\alpha(\nu \cond
\bmphi)$ and $\beta(\nu \cond \bmphi)$, we propose to start with a simple polynomial ansatz
\begin{align}
  \lleq{gmp}
  \alpha(\nu \cond \bmphi) &= \sum_{k=0}^\Kalpha \alpha_k \nu^k,
  &\beta(\nu \cond \bmphi) &= \sum_{k=0}^\Kbeta \beta_k \nu^k
\end{align}
such that
\begin{align}
  \lleq{gmq}
  \bmphi = (\alpha_0, \dots \alpha_\Kalpha , \beta_0, \dots \beta_\Kbeta)
\end{align}
The usual trade off between parsimony and the ability to accurately
model the frequency dependence determines the orders of the
polynomials. This is outside the scope of this paper.

\subsubsection*{Priors} \label{sec:priors}

To explicitly evaluate \refeq{gmd}, we need to specify the prior
$p(\bmphi)$. To gain some insight, we first consider the special case
$\alpha = \bmphi_1$, $\beta = \bmphi_2$ where we ignore the dependence
on $\nu$; i.e., $\Kalpha = \Kbeta = 0$. For this standard $\GammaDist$
model, there is a conjugate prior but unfortunately the normalization
constant cannot be evaluated in closed form, so it is thus of no use
for us. How about a noninformative prior? The authors of
\cite{moala2013bayesian} present an overview of possible choices and
conclude that the point estimate given by the mode of the posterior is
nearly insensitive to the prior choice starting from around 30
independent identically distributed samples. We posit that there are
this many samples available near any $\nu$ under consideration, so the
prior is not crucial for our application.

There is some explicit prior information that we want to build into
our analysis. The distribution $p(x \cond \alpha, \beta)$ should
always vanish at the origin. This is equivalent to requiring a maximum
away from the boundary leading us to the condition
\begin{align}
  \lleq{gmn}
  \alpha > 1.
\end{align}

For simplicity, we choose a uniform prior on $\bmphi$ on a finite box $V$ subject to the constraint \refeq{gmn}
\begin{align}
  p(\bmphi) = \mathbf{1}_{\alpha(\bmphi) > 1}(\bmphi) \mathbf{1}_V(\bmphi).
\end{align}
The box $V$ is chosen by trial and error to be large enough to contain
the relevant region of $p(\bmx\cond \bmphi, \npack, \bmnu)$. For
simple functions $\alpha(\nu \cond \bmphi)$, one can find the minimum
value of $\alpha$ over the entire range of $\nu$ analytically. If this
is too complicated, we propose to simplify by requiring $\alpha(\nu_j
\cond \phi) > 1$ only at each packet frequency $\nu_j$. In the
evaluation of $p(\bmx \cond \bmphi, \npack, \bmnu)$, one has to loop
over all packets and compute $\alpha$ and $\beta$ anyways, so the
constraint imposes essentially no computational overhead.

\subsection{Approximations}

To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
two difficulties: the sum over $N$ and the integral over
$\bmphi$. Translated to our problem, this would be 30 packets at the
\emph{same} frequency. But TARDIS only outputs packets at
\emph{different} frequencies. In the following, let $m$ denote the
effective number of packets close to $\nu$ where ``close'' is
deliberately left unspecified.  If $m$ is large, we can well determine
$\phi$ and thus $p(X_j \cond N, \bmphi, \nu)$.

@@

The following approximations can be used in different regions of the
parameter space spanned by $n$, $N$, and $m$. %
\hans{$n$ is TARDIS output ``data'', so not a parameter. From the
  above paragraph, $m$ seems to be a parameter governing rebinning to
  nonuniform bins. I would have no idea how to handle $m$ formally in
  a Bayesian way. If $m{=}150$ is the result of optimising the fitting
  stability, then presumably it will be kept constant. In that case,
  it would be treated like a constant rather than a parameter in the
  usual sense.}
\fred{Yes, $n$ is fixed from TARDIS output. $m$ is defined only for this paragraph}
\fred{Update approximations: don't need moments any longer}

\subsubsection*{Large $\bmn$} \label{large-n}

For large $n$, the cumulative of the Negative Binomial $p(N \cond n)$
can be approximated by the cumulative of a Gaussian with mean $n$ and
variance $2 (n - a + 1)$. Intuitively, this variance arises because the
Poisson variance appears twice, once when observing $n$ and once when
predicting $N$ given $n$.  The region around $N=n$ yields the dominant
contribution to the sum over $N$ in \refeq{gmd}. The contribution
around $N=0$ is then negligible, and we can approximate $\sum_N$ by
$\int \rmdx{N}$. The density $p(N \cond n) P(X \cond N, \npack, \bmx,
\bmnu)$ is close to a Gaussian (in the $x$ parametrization) and the
integral can be accurately approximated by Laplace's method which
requires the Hessian and the mode.
  % as each moment has to be
  % normalized by the evidence $p(\bmx \cond m)$.
The former is available in closed form and the latter has to be
determined numerically. Fortunately, Newton's method leads to a
solution in only a few steps~\cite{minka2002estimating}.
%
\fred{The reference is for max. likelihood of a $\GammaDist(\cdot)$
  but I'm sure the performance is similar for max. posterior and the
  first moments} We benefit from partial error cancelation when we
apply Laplace's method both in the numerator and
denominator~\cite{tierney_accurate_1986}.
  %
\fred{Might actually evaluate numerator and denominator at posterior
  mode to do only one instead of two optimizations;
  cf.~\cite{lindley_approximate_1980}}

\subsubsection*{Large $\bmN$} \label{large-N}

The central limit theorem states that for large $N$, $P(X \cond N,
\bmphi, \nu) \approx \mathcal{N}(X \cond N \, E_\bmphi[X_j \cond \nu],
N \, V_\bmphi[X_j \cond \nu])$. The expectation values have no known
closed-form solutions but we can easily and accurately approximate
them with Laplace's method.
% This is true independent of our model
assumption \refeq{gmi} that $X_j \sim \GammaDist(\alpha, \beta)$.
%
\fred{This only helps because the Laplace approx. when combined with
  large $n$ has an analytical result for the mode which saves us
  numerical optimization. But it requires extra (Laplace
  approximated?) integrals to compute $E_\bmphi[X_j \cond \nu]$ and
  $V_\bmphi[X_j \cond \nu]$. I can't get the Hessian analytically for
  $\npack$ samples. But we could use \texttt{Hesse} from
  \href{http://seal.web.cern.ch/seal/snapshot/work-packages/mathlibs/minuit/}{Minuit}
  or similar to approximate the Hessian at the mode by finite
  differences.}

\subsubsection*{Large $\bmm$}\label{large-m}

By asymptotic normality, the posterior $p(\bmphi \cond \npack, \bmx,
\bmnu) = p(\bmx\cond \bmphi, \npack, \bmnu) \, p(\bmphi) / p(\bmx
\cond \npack, \bmnu)$ approaches a multivariate Gaussian. In the limit
$m \to \infty$, the covariance of the posterior $p(\bmphi \cond
\npack, \bmx, \bmnu)$ tends to zero as $1/m$ and we can then fix the
parameters $\bmphi$ at the posterior mode $\bmphi^*$; i.e., $p(\bmphi
\cond \npack, \bmx, \bmnu) \to \delta(\bmphi - \bmphi^*)$ which removes the integral in \refeq{gmd}.

\subsubsection*{Large $\bmn, \bmN$,  and very large $\bmm$} \label{large-n-large-m}

With many samples observed in a narrow bin, both $n$ and $m$ are large
and the dominant contribution to \refeq{gmd} arises from large
$N$. Combining all of the above approximations, we find
\begin{align}
  \lleq{gmr} %
  p(X\cond \nu, \npack,\bmx,\bmnu) &= %
  \int \rmdx{N} \mathcal{N}(N \cond n, 2 (n -a + 1)) \, \mathcal{N}(X \cond
  N \, E_{\bmphi^*}[X_j \cond \nu], N \, V_{\bmphi^*}[X_j \cond \nu])
\end{align}
where
\begin{align}
  \lleq{gms}%
  E_{\bmphi^*}[X_j \cond \nu] &= \frac{\alpha(\nu \cond
    \bmphi^*)}{\beta(\nu \cond \bmphi^*)}, & %
  V_{\bmphi^*}[X_j \cond \nu] = \frac{\alpha(\nu \cond
    \bmphi^*)}{\beta(\nu \cond \bmphi^*)^2}.
\end{align}
Again, this integral does not have a closed-form expression but it is
unimodal and can be well approximated with Laplace's method. The mode
of the integrand is the single real root of a cubic polynomial. We
compute the root as well as the Hessian analytically with mathematica; the results are lengthy and not shown here.

\subsubsection*{Comparison of approximations}

@@
